From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-30 upi.tex
second derivation in 2.1

The $\AgdaFunction{not}$ function is a simple lifting of
\AgdaFunction{swap₊}. Using the
\AgdaFunction{controlled} building block, we can build a
controlled-not $\AgdaFunction{cnot}$ gate and a
controlled-controlled-not gate, also known as the
\AgdaFunction{toffoli} gate. The latter gate is a universal function
for combinational boolean circuits thus showing the expressiveness of
the language.

@@

{\small
\[\begin{array}{rcl}
\AgdaFunction{not} &:& \bt \leftrightarrow \bt \\
\AgdaFunction{not} &=& 
  \AgdaFunction{unfoldBool} \odot \AgdaFunction{swap₊} \odot \AgdaFunction{foldBool} \\
\\ 
\AgdaFunction{cnot} &:& \bt \otimes \bt \leftrightarrow \bt \otimes \bt \\
\AgdaFunction{cnot} &=& \AgdaFunction{controlled}~\AgdaFunction{not} \\
\\ 
\AgdaFunction{toffoli} &:& \bt \otimes (\bt \otimes \bt)
                           \leftrightarrow  \bt \otimes (\bt \otimes \bt) \\
\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot} \\
\end{array}\]}
%%%

The code for $\AgdaFunction{controlled}~\AgdaFunction{f}$ provides
constructive evidence (i.e., a program, a logic gate, or a hardware
circuit) for an automorphism on $\bt \otimes a$: it can be read
top-down or bottom-up to go back and forth. This perspective naturally
raises the question of when two programs are ``equivalent.'' Consider
the six programs of type $\bt \leftrightarrow \bt$ below:

\noindent We neither expect, nor desire, that all these programs of
the same type to be be equivalent. In the case above, each group of
three programs does represent ``the same'' program but the two groups
represent distinct programs. Thinking extensionally, i.e., by looking
at all possible input-output pairs, it is indeed easy to verify this
fact but we can do better. We can provide \emph{evidence} (i.e., a
program that manipulates programs) that can constructively transform
every program to an equivalent one. We show such a level-2 program
proving that $\AgdaFunction{not₃}$ is equivalent to
$\AgdaFunction{not}$. For illustration, the program for
$\AgdaFunction{not₃}$ is depicted in Fig.~\ref{fig:not}. We encourage
the reader to map the steps below to manipulations on the diagram that
would incrementally simplify it:

{\small 
\[\def\arraystretch{1.2}\begin{array}{rcl}
\AgdaFunction{notOpt} &:& \AgdaFunction{not₃} \Leftrightarrow \AgdaFunction{not} \\
\AgdaFunction{notOpt} &=& 
  \AgdaFunction{uniti⋆} \odot (\AgdaFunction{swap⋆} \odot 
                        ((\AgdaFunction{not} \otimes \AgdaFunction{id}) \odot 
                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}))) \\
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocLeft}} \\
&& \AgdaFunction{uniti⋆} \odot (\AgdaFunction{swap⋆} \odot 
                        (\AgdaFunction{not} \otimes \AgdaFunction{id})) \odot 
                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}) \\
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{swapLeft} 
                                  \boxdot \AgdaFunction{id})} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot \AgdaFunction{swap⋆}) \odot 
                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}) \\
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocRight}} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot (\AgdaFunction{swap⋆} \odot 
                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}))) \\
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
                                  \boxdot \AgdaFunction{assocLeft})} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot ((\AgdaFunction{swap⋆} \odot 
                      \AgdaFunction{swap⋆}) \odot \AgdaFunction{unite⋆})) \\
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
                                  \boxdot (\AgdaFunction{leftInv} \boxdot \AgdaFunction{id}))} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot (\AgdaFunction{id} \odot \AgdaFunction{unite⋆})) \\
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
                                  \boxdot \AgdaFunction{idLeft})} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot \AgdaFunction{unite⋆}) \\
&& \qquad\byisotwo{\AgdaFunction{assocLeft}} \\
&& (\AgdaFunction{uniti⋆} \odot (\AgdaFunction{id} \otimes \AgdaFunction{not})) 
                      \odot \AgdaFunction{unite⋆} \\
&& \qquad\byisotwo{\AgdaFunction{unitiLeft} \boxdot \AgdaFunction{id}} \\
&& (\AgdaFunction{not} \otimes \AgdaFunction{uniti⋆}) \odot \AgdaFunction{unite⋆} \\    
&& \qquad\byisotwo{\AgdaFunction{assocRight}} \\
&& \AgdaFunction{not} \otimes (\AgdaFunction{uniti⋆} \odot \AgdaFunction{unite⋆}) \\    
&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{leftInv}} \\
&& \AgdaFunction{not} \otimes \AgdaFunction{id} \\
&& \qquad\byisotwo{\AgdaFunction{idRight}} \\
&& \AgdaFunction{not} 
\end{array}\]}

>>>>

The $\AgdaFunction{not}$ function is a simple lifting of
\AgdaFunction{swap₊}. Using the
\AgdaFunction{controlled} building block, we can build a
controlled-not $\AgdaFunction{cnot}$ gate and a
controlled-controlled-not gate, also known as the
\AgdaFunction{toffoli} gate. The latter gate is a universal function
for combinational boolean circuits thus showing the expressiveness of
the language.
{\small
\[\begin{array}{rcl}
\AgdaFunction{not} &:& \bt \leftrightarrow \bt \\
\AgdaFunction{not} &=& 
  \AgdaFunction{unfoldBool} \odot \AgdaFunction{swap₊} \odot \AgdaFunction{foldBool} \\
\\ 
\AgdaFunction{cnot} &:& \bt \otimes \bt \leftrightarrow \bt \otimes \bt \\
\AgdaFunction{cnot} &=& \AgdaFunction{controlled}~\AgdaFunction{not} \\
\\ 
\AgdaFunction{toffoli} &:& \bt \otimes (\bt \otimes \bt)
                           \leftrightarrow  \bt \otimes (\bt \otimes \bt) \\
\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot} \\
\end{array}\]}
%%%

The code for $\AgdaFunction{controlled}~\AgdaFunction{f}$ provides
constructive evidence (i.e., a program, a logic gate, or a hardware
circuit) for an automorphism on $\bt \otimes a$: it can be read
top-down or bottom-up to go back and forth. This perspective naturally
raises the question of when two programs are ``equivalent.'' Consider
the follwing six programs of type~$\bt \leftrightarrow \bt$

@@

\noindent We neither expect, nor desire, that all these programs of
the same type to be be equivalent. In the above, each group of
three does represent ``the same'' program but the two groups
represent distinct programs. Thinking extensionally, i.e., by looking
at all possible input-output pairs, it is indeed easy to verify this
fact, but we can do better. We can provide \emph{evidence} (i.e., a
program that manipulates programs) that can constructively transform
every program to an equivalent one. We show such a level-2 program
proving that $\AgdaFunction{not₃}$ is equivalent to
$\AgdaFunction{not}$. For illustration, the program for
$\AgdaFunction{not₃}$ is depicted in Fig.~\ref{fig:not}. We encourage
the reader to map the steps below to manipulations on the diagram that
would incrementally simplify it:

{\small 
\[\def\arraystretch{1.2}\begin{array}{rcll}
\AgdaFunction{notOpt} &:& \AgdaFunction{not₃} \Leftrightarrow \AgdaFunction{not} \\
\AgdaFunction{notOpt} &=& 
  \AgdaFunction{uniti⋆} \odot (\AgdaFunction{swap⋆} \odot 
                        ((\AgdaFunction{not} \otimes \AgdaFunction{id}) \odot 
                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆})))
 & \quad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocLeft}} \\
&& \AgdaFunction{uniti⋆} \odot (\AgdaFunction{swap⋆} \odot 
                        (\AgdaFunction{not} \otimes \AgdaFunction{id})) \odot 
                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆})
 & \quad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{swapLeft} 
                                  \boxdot \AgdaFunction{id})} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot \AgdaFunction{swap⋆}) \odot 
                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆})
 & \quad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocRight}} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot (\AgdaFunction{swap⋆} \odot 
                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆})))
 & \quad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
                                  \boxdot \AgdaFunction{assocLeft})} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot ((\AgdaFunction{swap⋆} \odot 
                      \AgdaFunction{swap⋆}) \odot \AgdaFunction{unite⋆}))
 & \quad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
                                  \boxdot (\AgdaFunction{leftInv} \boxdot \AgdaFunction{id}))} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot (\AgdaFunction{id} \odot \AgdaFunction{unite⋆}))
 & \quad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
                                  \boxdot \AgdaFunction{idLeft})} \\
&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                      \odot \AgdaFunction{unite⋆})
 & \quad\byisotwo{\AgdaFunction{assocLeft}} \\
&& (\AgdaFunction{uniti⋆} \odot (\AgdaFunction{id} \otimes \AgdaFunction{not})) 
                      \odot \AgdaFunction{unite⋆}
 & \quad\byisotwo{\AgdaFunction{unitiLeft} \boxdot \AgdaFunction{id}} \\
&& (\AgdaFunction{not} \otimes \AgdaFunction{uniti⋆}) \odot \AgdaFunction{unite⋆}
 & \quad\byisotwo{\AgdaFunction{assocRight}} \\
&& \AgdaFunction{not} \otimes (\AgdaFunction{uniti⋆} \odot \AgdaFunction{unite⋆})
 & \quad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{leftInv}} \\
&& \AgdaFunction{not} \otimes \AgdaFunction{id}
 & \quad\byisotwo{\AgdaFunction{idRight}} \\
&& \AgdaFunction{not} 
\end{array}\]}
